### 2024-10-30 Data_Cleaning_20241030.Rmd
Create Data_Cleaning_20241030.Rmd
Data cleaning almost all done. Remaining tasks: (1) import spatial data and plot (2) come up with as many different ways to visualize the data as possible

---
title: "ECNS 560 Project - Bison Population"
author: "Kyle Froisland"
date: "2024-10-16"
output: html_document
---

Load Raw Data/Packages: *Data loaded from path on my local computer
```{r}
library(readr)
library(tidyverse)
library(readxl)
#Bison_numbers <- read_csv("MSU/ECNS 560/Term Project/Data/Bison_numbers_by_the_county National Bison Association.csv")
#hunting_data_tidy <- read_excel("MSU/GitHub/ECNS-560-Term-Project/Data/hunting_data_tidy.xlsx")
```

Create Tidy Version of Population Data:
```{r}
tidy_bison = Bison_numbers |>
  pivot_longer(cols=bison_census2002:bison_census2017, names_to="year", values_to="population") |>
  select(state_code, county, year, population)
```

Cleaning:
```{r}
#Change year Variable in tidy_bison to just the number (remove "bison_census")
tidy_bison_clean = mutate(tidy_bison, year = sapply(str_remove(tidy_bison$year, "bison_census"), function(x) (x[1])))

#define new dataset with by-state population vs by county
bison_data_merge = tidy_bison_clean |>
  group_by(state_code, year) |>
  summarize(total_population = sum(population)) |>
  rename_at(2, ~'year')

#Change column name of year in hunting data, filter to only years that match population data
colnames(hunting_data_tidy)[2] = 'year' 
hunting_data_tidy |>
  filter(year == "2002" | year == "2007" | year == "2012" | year == "2017")

#Ensure that years are numeric
bison_data_merge$year = as.numeric(bison_data_merge$year)
hunting_data_tidy$year = as.numeric(hunting_data_tidy$year)

```

Merge Data Sets:
```{r}
#join with hunting data by state and year as ID
merged_bison_data = left_join(bison_data_merge, hunting_data_tidy, by = c("state_code" = "state_code", "year" = "year"))
```

Load Map Data from Shapefile: 
```{r}
library(sf)
library(tidyverse)
library(tmap)
library(raster)
library(jsonlite)
library(rvest)
library(units)

#Load Habitats:
habitats = st_read("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/Bison Range Map/mBISOx_CONUS_Range_2001v1/mBISOx_CONUS_Range_2001v1.shp") |>
  st_make_valid()

#Calculate Area
st_area(habitats)

#Change Units from M^2 to KM^2
set_units(st_area(habitats), km^2)

plot(habitats)

#Load Ranges:
ranges = raster("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/USGS Bison Habitat Map/mBISOx_CONUS_HabMap_2001v1/mBISOx_CONUS_HabMap_2001v1.tif")

plot(ranges)

# Specify the path to your HTML file
file_path <- "C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/USGS Bison Habitat Map/mBISOx_CONUS_HabMap_2001v1/mBISOx_CONUS_HabMap_2001v1.tif.aux.xml"

# Read the HTML content
html_content <- read_html(file_path)

#Load JSON:
json = fromJSON("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/USGS Bison Habitat Map/mBISOx_CONUS_HabModel_2001v1.json")

```

Visualizations:
```{r}
graph1_data = merged_bison_data |>
  filter(state_code == "MONTANA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA")

ggplot(graph1_data, aes(x = year, y = total_population)) +
  geom_point() +
  labs(title = "Bison Population by State", x = "Year", y = "Bison Population") +
  facet_wrap(~state_code)
```

Next: (1) figure out how to load in the spacial data (2) need to play around with visualizations. If we both come up with as many different cool plots as we can we can pick and choose best ones.

### 2024-11-06 Data_Cleaning_20241031.Rmd
Draft Exploratory Analysis
See R Markdown file for draft exploratory analysis

---
title: "ECNS 560 Project - Bison Population"
author: "Kyle Froisland"
date: "2024-10-16"
output: html_document
---

Load Raw Data/Packages: *Data loaded from path on my local computer
```{r}
library(readr)
library(tidyverse)
library(readxl)
library(sf)
library(tmap)
library(raster)
library(rvest)
library(units)
library(dplyr)
library(ggplot2)

setwd("C:/Users/kdfro/Documents/MSU/ECNS 560/Term Project")
Bison_numbers <- read_csv("Data/Bison_numbers_by_the_county National Bison Association.csv")
hunting_data_tidy <- read_excel("Data/hunting_data_tidy.xlsx")
```

Create Tidy Version of Population Data:
```{r}
tidy_bison = Bison_numbers |>
  pivot_longer(cols=bison_census2002:bison_census2017, names_to="year", values_to="population") |>
  dplyr::select(state_code, county, year, population)
```

Cleaning:
```{r}
#Change year Variable in tidy_bison to just the number (remove "bison_census")
tidy_bison_clean = mutate(tidy_bison, year = sapply(str_remove(tidy_bison$year, "bison_census"), function(x) (x[1])))

#define new dataset with by-state population vs by county
bison_data_merge = tidy_bison_clean |>
  group_by(state_code, year) |>
  summarize(total_population = sum(population)) |>
  rename_at(2, ~'year')

#Change column name of year in hunting data, filter to only years that match population data
colnames(hunting_data_tidy)[2] = 'year' 
hunting_data_tidy |>
  filter(year == "2002" | year == "2007" | year == "2012" | year == "2017")

#Ensure that years are numeric
bison_data_merge$year = as.numeric(bison_data_merge$year)
hunting_data_tidy$year = as.numeric(hunting_data_tidy$year)

```

Merge Data Sets:
```{r}
#join with hunting data by state and year as ID
merged_bison_data = left_join(bison_data_merge, hunting_data_tidy, by = c("state_code" = "state_code", "year" = "year"))
```

Load Map Data from Shapefile: 
```{r}
setwd("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project")

#Load Habitats:
habitats = raster("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/USGS Bison Habitat Map/mBISOx_CONUS_HabMap_2001v1/mBISOx_CONUS_HabMap_2001v1.tif")

#Load Ranges:
ranges = st_read("Data/Bison Range Map/mBISOx_CONUS_Range_2001v1/mBISOx_CONUS_Range_2001v1.shp")

```

Quick Spatial Plot:
```{r}
qtm(ranges)
qtm(habitats)
```

Additional Visualizations:

```{r}
#Comparing population change over time across some Great Plains states
graph1_data = merged_bison_data |>
  filter(state_code == "MONTANA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA")

graph2_data = merged_bison_data |>
  filter(state_code == "MONTANA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA" | state_code == "NEBRASKA" | state_code == "KANSAS" | state_code == "OKLAHOMA" | state_code == "TEXAS" | state_code == "IDAHO" | state_code == "COLORADO" | state_code == "NEW MEXICO")

ggplot(graph1_data, aes(x = year, y = total_population)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Bison Population by State - Some Major Locations", x = "Year", y = "Bison Population") +
  facet_wrap(~state_code)

ggplot(graph2_data, aes(x = year, y = total_population)) + 
  xlab("Year") +
  ylab("Population") +
  labs(title = "Bison Population Over Time") +
  geom_point(aes(color = state_code), size = 3) +
  geom_smooth()

ggplot(graph2_data, aes(x = year, y = `Gross Cost - Hunting Licenses`)) + 
  xlab("Year") +
  ylab("Population") +
  labs(title = "Gross Cost of Hunting Licenses Over Time") +
  geom_point(aes(color = state_code), size = 3) +
  geom_smooth()

#Looking for relationship between number of hunting licenses and population
ggplot(merged_bison_data, aes(x = `Paid Hunting License Holders`, y = total_population)) +
  geom_point() + geom_smooth()
```

Saki's Plots:
```{r}
data = hunting_data_tidy

names(data)[2] <- "Year"

data <- data %>%
  mutate(Year = as.numeric(Year))

data_long <- data |>
  pivot_longer(
    cols = -c(Year, state_code),  # Exclude Year and state_code from pivoting
    names_to = "License_Type",     # New column to hold the names of original columns
    values_to = "Count"            # New column to hold the values
  )

# to line plot of each license type over time for a specific state (e.g., Montana)
ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
  geom_line() +
  geom_point() +
  labs(title = "Hunting License Trends Over Time in Montana", 
       x = "Year", 
       y = "Count / Cost",
       color = "License Type") +
  theme_minimal()

ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
  geom_line() +
  geom_point() +
  labs(title = "Hunting License Trends Over Time in Montana", 
       x = "Year", 
       y = "Count / Cost",
       color = "License Type") +
  theme_minimal() +
  facet_wrap(~ License_Type, scales = "free_y")  # To separate plots by License Type with individual y-scales

ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  labs(title = "Hunting License Trends Over Time in Montana", 
       x = "Year", 
       y = "Count / Cost",
       color = "License Type") +
  theme_minimal()

# Plot the data
ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
  geom_line() +
  geom_point() +
  scale_y_log10() +
  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
       x = "Year", 
       y = "Log Count / Cost",
       color = "License Type") +
  theme_minimal()

data_long$License_Type <- factor(data_long$License_Type, 
                                 labels = c("Cost Non-Resident", "Cost Resident", "Gross Cost", 
                                            "Non-Resident Licenses", "Paid Holders", "Resident Licenses"))

ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count)) +
  geom_line(aes(color = License_Type)) +
  geom_point(aes(color = License_Type)) +
  scale_y_log10() +
  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
       x = "Year", 
       y = "Log Count / Cost") +
  theme_minimal() +
  facet_wrap(~ License_Type, scales = "free_y")

ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess") +
  scale_y_log10() +
  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
       x = "Year", 
       y = "Log Count / Cost",
       color = "License Type") +
  theme_minimal()

# Load the scales package for custom label formatting
library(scales)

# Plot with custom y-axis labels
ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess") +
  scale_y_log10(labels = comma) +  # Use comma format for log scale labels
  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
       x = "Year", 
       y = "Count / Cost",
       color = "License Type") +
  theme_minimal()
```

### 2024-11-06 Draft Exploratory Analysis.Rmd
Draft Exploratory Analysis
See R Markdown file for draft exploratory analysis

---
title: "Draft Exploratory Analysis"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

***ECNS 560 Term Project: Bison Population Kyle Froisland and Saki***

***Topic:*** American Buffalo (Bison) Population Levels from Early 2000’s to
Now

***Potential Research Questions:*** 
1. How much has Bison population changed in the US over time? 
2. How much has Bison population changed in different parts of the US? 
3. How does Bison health vary across habitats/ranges? 
4. Do variations in state/national park laws and regulations impact Bison population levels? 
  a. Examples: availability of hunting tags, zoning/homesteading or prevalence of private ranches
under/over a certain size, regulations on meat/pelt sales (would need to locate data for specific “X” ideas) 
5. Do variations in state/national park laws and regulations impact average Bison population? 
6. How do changes in population impact hunting license spending?

***Motivation:*** Bison population reached near extinction (estimates
generally say a few hundred bison) during the late 19th century, but had
a miraculous comeback in the early to middle 20th century. However, the
health and sustainability of bison populations is still an issue.
Factors such as the opportunity cost of agricultural farming and cattle
ranching loom and tend to crowd out bison domestication, and policies
such as homesteading and urban expansion encroach on the grazing land of
bison. Bison, being the national mammal of the U.S., play a culturally
significant part in the history of the United States as well as the
culture of native populations prior to the establishment of the United
States. Thus, conservation efforts, regulations, or subsidies/incentives
for private herds and domestication may be worth consideration by policy
makers or conservation groups.

***Data:*** 
1. Bison population numbers by county from ArcGIS, Center of Excellence for Bison Studies, South Dakota State: Bison by the Numbers (arcgis.com) a. Years covered: 2002, 2007, 2012, 2017 b. Variables: State, County, Population (by year), Number of Operations (by year)

2.  United States Geological Survey (USGS) Bison Range and Bison Habitat
    Maps: American Bison (Bison bison) mBISOx_CONUS_2001v1 Range Map -
    ScienceBase-Catalog and American Bison (Bison bison)
    mBISOx_CONUS_2001v1 Habitat Map - ScienceBase-Catalog

<!-- -->

  a.  Provides spatial data on ranges and habitats

<!-- -->

3.  Hunting License Data By State:
    <https://us-east-1.quicksight.aws.amazon.com/sn/accounts/329180516311/dashboards/48b2aa9c-43a9-4ea6-887e-5465bd70140b?directory_alias=tracs-quicksight>

<!-- -->

  a.  Years Covered: 1958 - 2024
  b.  Variables: State, Year, Paid License Holders, Resident and
    Non-Resident licenses, Cost of Hunting Licenses

***Data Processing:***
  In order to analyze this data, we first created a tidy
version of the Bison Population data. In order to do this we pivoted the
data set from the columns containing population in each year and created
a "year" column and a "population" column in their place. Once this was
complete, we needed to string clean these columns as each observation
included "bison_census" before the year instead of containing only the
year. Then, we needed to create a new variable in order to get state-by
state data as opposed to county-by-county. While we lost some details
during this process, it allowed us to create a new dataset that would
merge cleanly with our hunting data (as the hunting data contains only
state level numbers).

  Next, we cleaned the hunting data, making sure that we rename columns so
that they will match with the population data and that we limited the
years to only those years that also match the population data (ie, 2002,
2007, 2012, and 2017). Additionally, some years were inputted as strings
across the data sets while others were numeric, so we converted all
years to numeric format in order to facilitate merging in the next step.

  Finally, we merged the two data sets by their state and year variables
in order to create a final dataframe for visualization and analysis. To
do this we used a left join method that would join the hunting data to
the original population data. The unit of analysis for this dataset is
population and hunting data by state and by year, which we chose because
we were constrained by the hunting data. Although we lost some details
from the population data, state and year is an appropriate level of
analysis because, as we see in the "ranges" shapefile, most Bison
populations are concentrated in herds that graze in very concentrated
and protected areas. Thus, additional details with regard to county
level population would not change our results significantly and can be
deduced from our range spatial data if necessary.

***Visualizations:***
  Our first visualization sought to explore the Bison population trends of some major states over time. The following displays trends among four major western United States states (Montana, Wyoming, North Dakota, and South Dakota):
![Major State Populations Over Time](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Major State Populations.png)

What we see from the initial visualization is that there seems to be a general downward trend in Bison population from 2002 to 2017, but that there is no obvious trend between sates in between. In other words, while all states ultimately have lower population numbers in 2017 than in 2002, there is fluctuation in between those years in which some states saw relatively steady decline (such as Montana), others saw sharp drops amidst periods of steady decline (such as North Dakota), and some even saw short periods of population growth (such as South Dakota and Wyoming).

  After realizing that there is some variety in the details of population growth or decline across states, we sought to confirm our suspician that the overall population trend is negative. What we see in the following graph is a confirmation of that suspicion, where population sees a strong downward trend across the Great Plains:
![Great Plains Population Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Population Over Time.png)

  Once we confirmed that population is indeed declining, we started exploring fluctuations in hunting license numbers as a possible correlate. The following graph shows the number of hunting licenses across the same years and the same states as our previous visual:
![Great Plains Hunting License Numbers Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Hunting Licenses Over Time.png)

From this graph, we do not see any obvious trends in hunting licenses over time across the Great Plains. The following graph confirms our suspician that there is not an obvious correlation between hunting licenses and Bison population overall (population fluctuates in both directions as the number of hunting licenses increases):
![Hunting License Numbers vs Population Over Time](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/License vs Population.png)

  Although there is not an obvious and immediate relationship with our initial visualizations, it is still possible that we could find a relationship if we can find quick changes in the number of licenses in certain areas to exploit. Thus, in the following graph we explored the number of licenses in various states to see if any saw rapid increases or decreases in hunting licenses:
![Hunting Licenses by State](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Hunting Licenses by State.png)

From this visualization, we see that most states do not have significant fluctuations in their number of hunting licenses. However, Texas saw an upward trend in licenses from 2012 to 2017. The following plot explores the relationship between the increase in hunting licenses and population changes in Texas:
![License vs Population in Texas](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/License vs Population in Texas.png)

The data does not seem to indicate a linear relationship. This could be due to a variety of reasons, including the possibility that an increase in the number of hunting licenses does not necessarily mean that there is an increase in the amount of Bison hunting (as the licenses could be for any number of animals).

  In further exploration, other indicators such as weather or local policy differences may be more useful indicators of population levels.

### 2024-11-06 Data_Cleaning_20241031.Rmd
Update to Draft Exploratory Analysis
Included spatial map as evidence for our decision to restrict analysis to Great Plains

@@
 
+graph3_data = merged_bison_data |>
+  filter(state_code == "TEXAS")
+
+#Some major state populations... all trends look different
 ggplot(graph1_data, aes(x = year, y = total_population)) +
@@
 
+#Overall trends across the great plains... see downward trend overall
 ggplot(graph2_data, aes(x = year, y = total_population)) + 
@@
 
-ggplot(graph2_data, aes(x = year, y = `Gross Cost - Hunting Licenses`)) + 
+#Number of Hunting licenses Over Time
+ggplot(graph2_data, aes(x = year, y = `Paid Hunting License Holders`)) + 
   xlab("Year") +
   ylab("Population") +
-  labs(title = "Gross Cost of Hunting Licenses Over Time") +
+  labs(title = "Number of Licenses Over Time") +
   geom_point(aes(color = state_code), size = 3) +
@@
 ggplot(merged_bison_data, aes(x = `Paid Hunting License Holders`, y = total_population)) +
-  geom_point() + geom_smooth()
+  geom_point() + 
+  geom_smooth() +
+  xlab("Number of Lisense Holders") +
+  ylab("Population") +
+  labs(title = "Number of Licenses vs Population")
 ```

### 2024-11-06 Draft Exploratory Analysis.Rmd
Update to Draft Exploratory Analysis
Included spatial map as evidence for our decision to restrict analysis to Great Plains

@@
 
-  After realizing that there is some variety in the details of population growth or decline across states, we sought to confirm our suspician that the overall population trend is negative.
+  After realizing that there is some variety in the details of population growth or decline across states, we sought to confirm our suspicion that the overall population trend is negative.
 What we see in the following graph is a confirmation of that suspicion, where population sees a strong downward trend across the Great Plains:
 ![Great Plains Population Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Population Over Time.png)
+
+  As a side note, our decision to restrict our focus to the Great Plains was based off of our spatial data which indicates to us that the majority of Bison herds are concentrated in these states, as shows in the following image:
+![Bison Herds in the U.S.](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Range Map.png)
 

### 2024-11-07 Data_Cleaning_20241031.Rmd
Final Version of Draft Exploratory Analysis
Fixed Y label for license over time graph

@@
 #Load Habitats:
-habitats = raster("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/USGS Bison Habitat Map/mBISOx_CONUS_HabMap_2001v1/mBISOx_CONUS_HabMap_2001v1.tif")
+#habitats = raster("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/USGS Bison Habitat Map/mBISOx_CONUS_HabMap_2001v1/mBISOx_CONUS_HabMap_2001v1.tif")
 

### 2024-11-07 Draft Exploratory Analysis.Rmd
Final Version of Draft Exploratory Analysis
Fixed Y label for license over time graph

@@
 The following graph shows the number of hunting licenses across the same years and the same states as our previous visual:
-![Great Plains Hunting License Numbers Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Hunting Licenses Over Time.png)
+![Great Plains Hunting License Numbers Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Licenses Over Time Fixed.png)
 

### 2024-12-02 Data_Cleaning_20241031.Rmd
Add files via upload

--- 2024-11-06 Data_Cleaning_20241031.Rmd
@@
 library(ggplot2)
+library(units)
 
-setwd("C:/Users/kdfro/Documents/MSU/ECNS 560/Term Project")
-Bison_numbers <- read_csv("Data/Bison_numbers_by_the_county National Bison Association.csv")
-hunting_data_tidy <- read_excel("Data/hunting_data_tidy.xlsx")
+Bison_numbers <- read_csv("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/Bison_numbers_by_the_county National Bison Association.csv")
+hunting_data_tidy <- read_excel("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/hunting_data_tidy.xlsx")
 ```
@@
 
-#Load Habitats:
-habitats = raster("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/USGS Bison Habitat Map/mBISOx_CONUS_HabMap_2001v1/mBISOx_CONUS_HabMap_2001v1.tif")
-
 #Load Ranges:
@@
 
+#Load States:
+states = st_read("C:/Users/kdfro/Documents/MSU/GitHub/course-materials/lecture-slides/12-Spatial/data/us_states_contiguous/states_contiguous.shp")
+```
+Merge States and Ranges:
+```{r}
+#Check CRS
+st_crs(states)
+st_crs(ranges)
+
+#Transform/Project:
+ranges2 = st_transform(ranges, crs = st_crs(states))
+stopifnot(st_crs(ranges2) == (st_crs(states)))
+
+tm_shape(states) + tm_polygons() +
+  tm_shape(ranges2) + tm_polygons(col = "darkgreen")
 ```
 
-Quick Spatial Plot:
+Spatial Cleaning/Exploration:
 ```{r}
-qtm(ranges)
-qtm(habitats)
+#Split ranges df into multiple polygons
+ranges3 <- st_cast(ranges2, "POLYGON")
+
+#Create variable for area
+ranges3$area = st_area(ranges3)
+
+#Change units from m^2 to km^2
+set_units(st_area(ranges3), km^2)
+
+#Visualize
+tm_shape(ranges3) + tm_polygons(col = "area")
+
+#Select states of interest:
+states2 = states |>
+  filter(STATE == "Colorado" | STATE == "Idaho" | STATE == "Kansas" | STATE == "Montana" | STATE == "Nebraska" | STATE == "New Mexico" | STATE == "North Dakota" | STATE == "North Dakota" | STATE == "Oklahoma" | STATE == "South Dakota" | STATE == "Texas" | STATE == "Wyoming")
+
+#Find intersection of ranges and states:
+ranges_intersect = st_filter(ranges3, states2)
+
+#Visualize:
+tm_shape(states2) + tm_polygons() +
+  tm_shape(ranges_intersect) + tm_polygons(col = "darkgreen")
 ```
@@
 graph2_data = merged_bison_data |>
-  filter(state_code == "MONTANA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA" | state_code == "NEBRASKA" | state_code == "KANSAS" | state_code == "OKLAHOMA" | state_code == "TEXAS" | state_code == "IDAHO" | state_code == "COLORADO" | state_code == "NEW MEXICO")
+  filter(state_code == "MONTANA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA" | state_code == "NEBRASKA" | state_code == "KANSAS" | state_code == "OKLAHOMA" | state_code == "IDAHO" | state_code == "COLORADO" | state_code == "NEW MEXICO") |>
+  mutate(population_bin = cut(total_population, breaks = c(0, 2000, 5000, 10000), labels = c("0-2000", "2000-5000", "5000-10000")))
 
@@
   filter(state_code == "TEXAS")
+
+graph4_data = merged_bison_data |>
+  filter(state_code == "OKLAHOMA")
 
@@
   xlab("Year") +
-  ylab("Population") +
+  ylab("Population by State") +
   labs(title = "Bison Population Over Time") +
-  geom_point(aes(color = state_code), size = 3) +
+  geom_point(aes(color = population_bin), size = 3) +
   geom_smooth()
+#Each point on this plot represents a state for a given year, and the color represents the population bin that the state falls into.
 
@@
   xlab("Year") +
-  ylab("Population") +
+  ylab("Number of Licenses") +
   labs(title = "Number of Licenses Over Time") +
-  geom_point(aes(color = state_code), size = 3) +
-  geom_smooth()
+  geom_point(aes(color = state_code), size = 3)
+#Removed Texas from this plot because it was an extreme value and make the graph less interpretable by extending the y-axis so much.
+Texas saw a small increase over the last time period but was otherwise steady.
 
@@
   xlab("Number of Lisense Holders") +
-  ylab("Population") +
+  ylab("Bison Population") +
   labs(title = "Number of Licenses vs Population")
+#Not a clear relationship. With so many states there are too many competing factors at play to control for with our limited data.
+
+#What about just Oklahoma since it is one of the only states with increasing numbers of licenses?
+ggplot(graph4_data, aes(x = `Paid Hunting License Holders`, y = total_population)) +
+  geom_point() + 
+  xlab("Number of Lisense Holders") +
+  ylab("Bison Population") +
+  labs(title = "Number of Licenses vs Population for Oklahoma")
+#We still have not inferred causality, but there seems to be a negative relationship now.
+The lack of variation in hunting licenses across other states makes regression extremely difficult, but Oklahoma could be a better case study.
+
 ```

### 2024-12-02 Draft Exploratory Analysis.Rmd
Add files via upload

@@
 What we see in the following graph is a confirmation of that suspicion, where population sees a strong downward trend across the Great Plains:
-![Great Plains Population Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Population Over Time.png)
+![Great Plains Population Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/bison pop over time by state and binned.png)
+In this plot, each point on this plot represents a state for a given year, and the color represents the population bin that the state falls into.
 
   As a side note, our decision to restrict our focus to the Great Plains was based off of our spatial data which indicates to us that the majority of Bison herds are concentrated in these states, as shows in the following image:
-![Bison Herds in the U.S.](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Range Map.png)
+![Bison Herds in the U.S.](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/States of interest and ranges map.png)
 
@@
 The following graph shows the number of hunting licenses across the same years and the same states as our previous visual:
-![Great Plains Hunting License Numbers Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Licenses Over Time Fixed.png)
+![Great Plains Hunting License Numbers Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/licenses over time 2.png)
+Note that we removed texas from this graph.
+It was an extreme value which extended our y-axis greatly.
+We noted that Texas had some variation from 2010 to 2017 which is useful for us, however it was excluded from the graph in order to better demonstrate variation in other states and make the plots more interpretable.
 
-From this graph, we do not see any obvious trends in hunting licenses over time across the Great Plains.
-The following graph confirms our suspician that there is not an obvious correlation between hunting licenses and Bison population overall (population fluctuates in both directions as the number of hunting licenses increases):
-![Hunting License Numbers vs Population Over Time](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/License vs Population.png)
+  From this graph, we do not see any obvious trends in hunting licenses over time across most states in the Great Plains.
+The following graph confirms our suspicion that there is not an obvious correlation between hunting licenses and Bison population overall (population fluctuates in both directions as the number of hunting licenses increases):
+![Hunting License Numbers vs Population Over Time](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/license vs population 2.png)
 
@@
 
-From this visualization, we see that most states do not have significant fluctuations in their number of hunting licenses.
+From this visualization, we see that most states do not have significant variation in their number of hunting licenses.
 However, Texas saw an upward trend in licenses from 2012 to 2017.
@@
 
+  The other state that saw some variation in the number of hunting licenses issued is Oklahoma, as shown below:
+![License vs Population in Texas](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Oklahoma Hunting vs Pop.png)
+Although we still lack sufficient evidence to infer causality between the two variables, Oklahoma at least demonstrates a negative relationship between hunting licenses and Bison population.
+
+  We initially expected a negative relationship between hunting licenses and bison populations as a consequence of overhunting a common pool resource across all states.
+If states were to issue more licenses, we thought that this would ultimately mean that populations decrease as a consequence of more bison being killed by hunters.
+We believe that this relationship is not so obvious for two main reasons.
+First, the allocation of hunting licenses is heavily regulated at the state level in order to prevent a tragedy of the commons.
+Second, a state's capacity to maintain large Bison populations is dependent on the amount of land that the animals have to graze.
+Thus, an increase in the number of hunting licenses issued could in fact be an indicator of an increased amount of private hunting land.
+As the demand for Bison tags increases, it raises the opportunity cost of ranchers with land adjacent to bison ranges and incentives them to allow Bison to expand their territory.
+Thus, although more Bison are being hunted, there is more land available for additional growth.
+
   In further exploration, other indicators such as weather or local policy differences may be more useful indicators of population levels.

### 2024-12-02 Data_Cleaning_20241031.Rmd
Add files via upload

@@
+<<<<<<< HEAD
+---
+title: "ECNS 560 Project - Bison Population"
+author: "Kyle Froisland"
+date: "2024-10-16"
+output: html_document
+---
+
+Load Raw Data/Packages: *Data loaded from path on my local computer
+```{r}
+library(readr)
+library(tidyverse)
+library(readxl)
+library(sf)
+library(tmap)
+library(raster)
+library(rvest)
+library(units)
+library(dplyr)
+library(ggplot2)
+library(units)
+
+Bison_numbers <- read_csv("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/Bison_numbers_by_the_county National Bison Association.csv")
+hunting_data_tidy <- read_excel("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/hunting_data_tidy.xlsx")
+```
+
+Create Tidy Version of Population Data:
+```{r}
+tidy_bison = Bison_numbers |>
+  pivot_longer(cols=bison_census2002:bison_census2017, names_to="year", values_to="population") |>
+  dplyr::select(state_code, county, year, population)
+```
+
+Cleaning:
+```{r}
+#Change year Variable in tidy_bison to just the number (remove "bison_census")
+tidy_bison_clean = mutate(tidy_bison, year = sapply(str_remove(tidy_bison$year, "bison_census"), function(x) (x[1])))
+
+#define new dataset with by-state population vs by county
+bison_data_merge = tidy_bison_clean |>
+  group_by(state_code, year) |>
+  summarize(total_population = sum(population)) |>
+  rename_at(2, ~'year')
+
+#Change column name of year in hunting data, filter to only years that match population data
+colnames(hunting_data_tidy)[2] = 'year' 
+hunting_data_tidy |>
+  filter(year == "2002" | year == "2007" | year == "2012" | year == "2017")
+
+#Ensure that years are numeric
+bison_data_merge$year = as.numeric(bison_data_merge$year)
+hunting_data_tidy$year = as.numeric(hunting_data_tidy$year)
+
+```
+
+Merge Data Sets:
+```{r}
+#join with hunting data by state and year as ID
+merged_bison_data = left_join(bison_data_merge, hunting_data_tidy, by = c("state_code" = "state_code", "year" = "year"))
+```
+
+Load Map Data from Shapefile: 
+```{r}
+setwd("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project")
+
+#Load Ranges:
+ranges = st_read("Data/Bison Range Map/mBISOx_CONUS_Range_2001v1/mBISOx_CONUS_Range_2001v1.shp")
+
+#Load States:
+states = st_read("C:/Users/kdfro/Documents/MSU/GitHub/course-materials/lecture-slides/12-Spatial/data/us_states_contiguous/states_contiguous.shp")
+```
+Fix States and Ranges:
+```{r}
+#Check CRS
+st_crs(states)
+st_crs(ranges)
+
+#Transform/Project:
+ranges2 = st_transform(ranges, crs = st_crs(states))
+stopifnot(st_crs(ranges2) == (st_crs(states)))
+
+tm_shape(states) + tm_polygons() +
+  tm_shape(ranges2) + tm_polygons(col = "darkgreen")
+```
+
+Spatial Cleaning/Exploration:
+```{r}
+#Split ranges df into multiple polygons
+ranges3 <- st_cast(ranges2, "POLYGON")
+
+#Create variable for area
+ranges3$area = st_area(ranges3)
+
+#Change units from m^2 to km^2
+set_units(st_area(ranges3), km^2)
+
+#Visualize
+tm_shape(ranges3) + tm_polygons(col = "area")
+
+#Select states of interest:
+states2 = states |>
+  filter(STATE == "Colorado" | STATE == "Idaho" | STATE == "Kansas" | STATE == "Montana" | STATE == "Nebraska" | STATE == "New Mexico" | STATE == "North Dakota" | STATE == "North Dakota" | STATE == "Oklahoma" | STATE == "South Dakota" | STATE == "Texas" | STATE == "Wyoming")
+
+#Find intersection of ranges and states:
+ranges_intersect = st_filter(ranges3, states2)
+
+#Visualize:
+tm_shape(states2) + tm_polygons() +
+  tm_shape(ranges_intersect) + tm_polygons(col = "darkgreen")
+```
+
+Additional Visualizations:
+
+```{r}
+#Comparing population change over time across some Great Plains states
+graph1_data = merged_bison_data |>
+  filter(state_code == "MONTANA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA")
+
+graph2_data = merged_bison_data |>
+  filter(state_code == "MONTANA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA" | state_code == "NEBRASKA" | state_code == "KANSAS" | state_code == "OKLAHOMA" | state_code == "IDAHO" | state_code == "COLORADO" | state_code == "NEW MEXICO") |>
+  mutate(population_bin = cut(total_population, breaks = c(0, 2000, 5000, 10000), labels = c("0-2000", "2000-5000", "5000-10000")))
+
+graph3_data = merged_bison_data |>
+  filter(state_code == "TEXAS")
+
+graph4_data = merged_bison_data |>
+  filter(state_code == "OKLAHOMA")
+
+graph5_data = merged_bison_data |>
+  filter(state_code == "MONTANA" | state_code == "TEXAS" | state_code == "OKLAHOMA" | state_code == "WYOMING" | state_code == "NORTH DAKOTA" | state_code == "SOUTH DAKOTA" | state_code == "NEBRASKA" | state_code == "KANSAS" | state_code == "OKLAHOMA" | state_code == "IDAHO" | state_code == "COLORADO" | state_code == "NEW MEXICO")
+
+#Some major state populations... all trends look different
+ggplot(graph1_data, aes(x = year, y = total_population)) +
+  geom_point() +
+  geom_smooth() +
+  labs(title = "Bison Population by State - Some Major Locations", x = "Year", y = "Bison Population") +
+  facet_wrap(~state_code)
+
+#Populations for all states of interest. There seems to be larger decreases in population around western planes states than southwestern states.
+ggplot(graph5_data, aes(x = year, y = total_population)) +
+  geom_point() +
+  geom_smooth() +
+  labs(title = "Bison Population by State - All States of Interest", x = "Year", y = "Bison Population") +
+  facet_wrap(~state_code)
+#This is also biased because southwestern states started with much lower populations to begin with.
+
+#Overall trends across the great plains... see downward trend overall
+ggplot(graph2_data, aes(x = year, y = total_population)) + 
+  xlab("Year") +
+  ylab("Population by State") +
+  labs(title = "Bison Population Over Time") +
+  geom_point(aes(color = population_bin), size = 3) +
+  geom_smooth()
+#Each point on this plot represents a state for a given year, and the color represents the population bin that the state falls into.
+
+#Number of Hunting licenses Over Time
+ggplot(graph2_data, aes(x = year, y = `Paid Hunting License Holders`)) + 
+  xlab("Year") +
+  ylab("Number of Licenses") +
+  labs(title = "Number of Licenses Over Time") +
+  geom_point(aes(color = state_code), size = 3)
+#Removed Texas from this plot because it was an extreme value and make the graph less interpretable by extending the y-axis so much.
+Texas saw a small increase over the last time period but was otherwise steady.
+
+#Looking for relationship between number of hunting licenses and population
+ggplot(merged_bison_data, aes(x = `Paid Hunting License Holders`, y = total_population)) +
+  geom_point() + 
+  geom_smooth() +
+  xlab("Number of Lisense Holders") +
+  ylab("Bison Population") +
+  labs(title = "Number of Licenses vs Population")
+#Not a clear relationship. With so many states there are too many competing factors at play to control for with our limited data.
+
+#What about just Oklahoma since it is one of the only states with increasing numbers of licenses?
+ggplot(graph4_data, aes(x = `Paid Hunting License Holders`, y = total_population)) +
+  geom_point() + 
+  xlab("Number of Lisense Holders") +
+  ylab("Bison Population") +
+  labs(title = "Number of Licenses vs Population for Oklahoma")
+#We still have not inferred causality, but there seems to be a negative relationship now.
+The lack of variation in hunting licenses across other states makes regression extremely difficult, but Oklahoma could be a better case study.
+
+```
+
+Research Question 1: Can we predict percent decrease in population levels based on Latitude of their centroid (Proxied by proximity to Saskatchewan)?
+```{r}
+#Load Canada Shapefile:
+Canada = st_read("C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Data/ca_shp/ca.shp")
+
+#Convert GRS:
+Canada2 = st_transform(Canada, crs = st_crs(states2))
+stopifnot(st_crs(Canada2) == (st_crs(states2)))
+
+#Find Centroids of states of interest:
+state_centroids = st_centroid(states2)
+
+#Plot to confirm:
+tm_shape(states2) + tm_borders() +
+  tm_shape(states2) + tm_dots(col = "red")
+
+#Make a variable for proximity of centroid to Saskatchewan:
+distance_pop_data = state_centroids |>
+  mutate(distance = st_distance(state_centroids$geometry, Canada2[3,4]))
+
+#Merge data with population, create variables for distance from Canada and population decline:
+distance = as.data.frame(distance_pop_data) |>
+  rename(state_code = STATE) |>
+  mutate(state_code = toupper(state_code)) |>
+  left_join(bison_data_merge, by = "state_code") |>
+  group_by(state_code) |>
+  mutate(difference = abs(total_population - total_population[year == 2002])/total_population[year == 2002]) |>
+  filter(year == 2017)
+
+#Plot Proximity vs Population Decrease:
+ggplot(distance, aes(x = log(distance), y = difference)) + 
+  xlab("log(distance) from Canada") +
+  ylab("Percent Decrease in Bison Population") +
+  labs(title = "Latitude and Bison Population Decline") +
+  geom_point()
+#There does not seem to be as clear of a relationship from this plot as I had originally expected. The predictive power of this variable seems tenuous.
+```
+
+Research Question 2: Can we predict population decline using Latitude and hunting data? This probably won't work due to limited data points...
+```{r}
+library(tidyverse)
+library(tidymodels)
+library(ISLR)
+library(glmnet)
+
+#Merge Distance data set with hunting data:
+distance_hunting = left_join(distance, hunting_data_tidy, by = c("state_code" = "state_code", "year" = "year")) |> as.data.frame()
+
+#Create Predictive Model:
+#First, split into train and test data
+set.seed(123)
+bison_split = distance_hunting |> initial_split(prop = 0.7)
+bison_train = distance_hunting |> training()
+bison_test = distance_hunting |> testing()
+
+#Define a Recipe:
+bison_recipe = recipe(difference ~ disatnce + total_population + Paid Hunting License Holders, data = bison_train) |> 
+  update_role(GEOID, new_role = "id") |> 
+  step_poly(all_numeric_predictors(), degree = 2) |> #create second term for each numeric var that is a quadratic
+  step_interact(~all_numeric_predictors():all_nominal_predictors()) |> #create interactions between numeric and indicator variables
+  step_dummy(all_nominal_predictors()) |> #create indicators for each categorical or character variable
+  step_nzv(all_predictors())#remove predictors with nearly 0 variance... otherwise not very reliable because difficult to pickup patterns
+
+#Process the Data:
+bison_clean = bison_recipe |> prep() |> juice()
+
+#Define linear model:
+model_lm = linear_reg() |> set_engine("lm")
+
+#Define cross validation for experimentation with other regression forms later
+bison_cv = bison_train |> vfold_cv(v=5)
+
+#Define Workflow:
+fit_lm_cv = workflow() |>
+  add_model(model_lm) |>
+  add_recipe(bisont_recipe) |>
+  fit_resamples(bison_cv)
+
+#Fit Model:
+fit_lm_cv |> collect_metrics(summarize = F)
+
+```
+
+Saki's Plots:
+```{r}
+data = hunting_data_tidy
+
+names(data)[2] <- "Year"
+
+data <- data %>%
+  mutate(Year = as.numeric(Year))
+
+data_long <- data |>
+  pivot_longer(
+    cols = -c(Year, state_code),  # Exclude Year and state_code from pivoting
+    names_to = "License_Type",     # New column to hold the names of original columns
+    values_to = "Count"            # New column to hold the values
+  )
+
+# to line plot of each license type over time for a specific state (e.g., Montana)
+ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
+  geom_line() +
+  geom_point() +
+  labs(title = "Hunting License Trends Over Time in Montana", 
+       x = "Year", 
+       y = "Count / Cost",
+       color = "License Type") +
+  theme_minimal()
+
+ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
+  geom_line() +
+  geom_point() +
+  labs(title = "Hunting License Trends Over Time in Montana", 
+       x = "Year", 
+       y = "Count / Cost",
+       color = "License Type") +
+  theme_minimal() +
+  facet_wrap(~ License_Type, scales = "free_y")  # To separate plots by License Type with individual y-scales
+
+ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
+  geom_point() +
+  geom_smooth(se = FALSE) +
+  labs(title = "Hunting License Trends Over Time in Montana", 
+       x = "Year", 
+       y = "Count / Cost",
+       color = "License Type") +
+  theme_minimal()
+
+# Plot the data
+ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
+  geom_line() +
+  geom_point() +
+  scale_y_log10() +
+  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
+       x = "Year", 
+       y = "Log Count / Cost",
+       color = "License Type") +
+  theme_minimal()
+
+data_long$License_Type <- factor(data_long$License_Type, 
+                                 labels = c("Cost Non-Resident", "Cost Resident", "Gross Cost", 
+                                            "Non-Resident Licenses", "Paid Holders", "Resident Licenses"))
+
+ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count)) +
+  geom_line(aes(color = License_Type)) +
+  geom_point(aes(color = License_Type)) +
+  scale_y_log10() +
+  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
+       x = "Year", 
+       y = "Log Count / Cost") +
+  theme_minimal() +
+  facet_wrap(~ License_Type, scales = "free_y")
+
+ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
+  geom_point() +
+  geom_smooth(se = FALSE, method = "loess") +
+  scale_y_log10() +
+  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
+       x = "Year", 
+       y = "Log Count / Cost",
+       color = "License Type") +
+  theme_minimal()
+
+# Load the scales package for custom label formatting
+library(scales)
+
+# Plot with custom y-axis labels
+ggplot(data_long %>% filter(state_code == "MONTANA"), aes(x = Year, y = Count, color = License_Type)) +
+  geom_point() +
+  geom_smooth(se = FALSE, method = "loess") +
+  scale_y_log10(labels = comma) +  # Use comma format for log scale labels
+  labs(title = "Hunting License Trends Over Time in Montana (Log Scale)", 
+       x = "Year", 
+       y = "Count / Cost",
+       color = "License Type") +
+  theme_minimal()
+```
+=======
 ---
@@
 ```
+>>>>>>> ae5d3f30ef6b3c153d8c3a489f3f9dc3d1e461cb

### 2024-12-02 Draft Exploratory Analysis.Rmd
Add files via upload

@@
+<<<<<<< HEAD
+---
+title: "Draft Exploratory Analysis"
+output: html_document
+editor_options: 
+  markdown: 
+    wrap: 72
+---
+
+***ECNS 560 Term Project: Bison Population Kyle Froisland and Saki***
+
+***Topic:*** American Buffalo (Bison) Population Levels from Early 2000’s to
+Now
+
+***Potential Research Questions:*** 
+1. How much has Bison population changed in the US over time? 
+2. How much has Bison population changed in different parts of the US? 
+3. How does Bison health vary across habitats/ranges? 
+4. Do variations in state/national park laws and regulations impact Bison population levels? 
+  a. Examples: availability of hunting tags, zoning/homesteading or prevalence of private ranches
+under/over a certain size, regulations on meat/pelt sales (would need to locate data for specific “X” ideas) 
+5. Do variations in state/national park laws and regulations impact average Bison population? 
+6. How do changes in population impact hunting license spending?
+
+***Motivation:*** Bison population reached near extinction (estimates
+generally say a few hundred bison) during the late 19th century, but had
+a miraculous comeback in the early to middle 20th century. However, the
+health and sustainability of bison populations is still an issue.
+Factors such as the opportunity cost of agricultural farming and cattle
+ranching loom and tend to crowd out bison domestication, and policies
+such as homesteading and urban expansion encroach on the grazing land of
+bison. Bison, being the national mammal of the U.S., play a culturally
+significant part in the history of the United States as well as the
+culture of native populations prior to the establishment of the United
+States. Thus, conservation efforts, regulations, or subsidies/incentives
+for private herds and domestication may be worth consideration by policy
+makers or conservation groups.
+
+***Data:*** 
+1.
+Bison population numbers by county from ArcGIS, Center of Excellence for Bison Studies, South Dakota State: Bison by the Numbers (arcgis.com) a.
+Years covered: 2002, 2007, 2012, 2017 b.
+Variables: State, County, Population (by year), Number of Operations (by year)
+
+2.  United States Geological Survey (USGS) Bison Range and Bison Habitat
+    Maps: American Bison (Bison bison) mBISOx_CONUS_2001v1 Range Map -
+    ScienceBase-Catalog and American Bison (Bison bison)
+    mBISOx_CONUS_2001v1 Habitat Map - ScienceBase-Catalog
+
+<!-- -->
+
+  a.  Provides spatial data on ranges and habitats
+
+<!-- -->
+
+3.  Hunting License Data By State:
+    <https://us-east-1.quicksight.aws.amazon.com/sn/accounts/329180516311/dashboards/48b2aa9c-43a9-4ea6-887e-5465bd70140b?directory_alias=tracs-quicksight>
+
+<!-- -->
+
+  a.  Years Covered: 1958 - 2024
+  b.  Variables: State, Year, Paid License Holders, Resident and
+    Non-Resident licenses, Cost of Hunting Licenses
+
+***Data Processing:***
+  In order to analyze this data, we first created a tidy
+version of the Bison Population data. In order to do this we pivoted the
+data set from the columns containing population in each year and created
+a "year" column and a "population" column in their place. Once this was
+complete, we needed to string clean these columns as each observation
+included "bison_census" before the year instead of containing only the
+year. Then, we needed to create a new variable in order to get state-by
+state data as opposed to county-by-county. While we lost some details
+during this process, it allowed us to create a new dataset that would
+merge cleanly with our hunting data (as the hunting data contains only
+state level numbers).
+
+  Next, we cleaned the hunting data, making sure that we rename columns so
+that they will match with the population data and that we limited the
+years to only those years that also match the population data (ie, 2002,
+2007, 2012, and 2017). Additionally, some years were inputted as strings
+across the data sets while others were numeric, so we converted all
+years to numeric format in order to facilitate merging in the next step.
+
+  Finally, we merged the two data sets by their state and year variables
+in order to create a final dataframe for visualization and analysis. To
+do this we used a left join method that would join the hunting data to
+the original population data. The unit of analysis for this dataset is
+population and hunting data by state and by year, which we chose because
+we were constrained by the hunting data. Although we lost some details
+from the population data, state and year is an appropriate level of
+analysis because, as we see in the "ranges" shapefile, most Bison
+populations are concentrated in herds that graze in very concentrated
+and protected areas. Thus, additional details with regard to county
+level population would not change our results significantly and can be
+deduced from our range spatial data if necessary.
+
+***Visualizations:***
+  Our first visualization sought to explore the Bison population trends of some major states over time.
+The following displays trends among four major western United States states (Montana, Wyoming, North Dakota, and South Dakota):
+![Major State Populations Over Time](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Major State Populations.png)
+
+What we see from the initial visualization is that there seems to be a general downward trend in Bison population from 2002 to 2017, but that there is no obvious trend between sates in between.
+In other words, while all states ultimately have lower population numbers in 2017 than in 2002, there is fluctuation in between those years in which some states saw relatively steady decline (such as Montana), others saw sharp drops amid periods of steady decline (such as North Dakota), and some even saw short periods of population growth (such as South Dakota and Wyoming).
+
+  After realizing that there is some variety in the details of population growth or decline across states, we sought to confirm our suspicion that the overall population trend is negative.
+What we see in the following graph is a confirmation of that suspicion, where population sees a strong downward trend across the Great Plains:
+![Great Plains Population Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/bison pop over time by state and binned.png)
+In this plot, each point on this plot represents a state for a given year, and the color represents the population bin that the state falls into.
+We can see that the number of blue (high population) and green (medium population) observations decreases over time, as the number of orange (low population) observations increases over time.
+This shows that states are generally trending down from higher population bins to lower population bins and this general trend is captured by the downward sloping local regression line.
+
+  As a side note, our decision to restrict our focus to the Great Plains was based off of our spatial data which indicates to us that the majority of Bison herds are concentrated in these states, as shows in the following image:
+![Bison Herds in the U.S.](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/States of interest and ranges map.png)
+
+  Once we confirmed that population is indeed declining, we started exploring fluctuations in hunting license numbers as a possible correlate.
+The following graph shows the number of hunting licenses across the same years and the same states as our previous visual:
+![Great Plains Hunting License Numbers Over Time - Overall Trend](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/licenses over time 2.png)
+Note that we removed texas from this graph.
+It was an extreme value which extended our y-axis greatly.
+We noted that Texas had some variation from 2010 to 2017 which is useful for us, however it was excluded from the graph in order to better demonstrate variation in other states and make the plots more interpretable.
+
+  From this graph, we do not see any obvious trends in hunting licenses over time across most states in the Great Plains.
+The following graph confirms our suspicion that there is not an obvious correlation between hunting licenses and Bison population overall (population fluctuates in both directions as the number of hunting licenses increases):
+![Hunting License Numbers vs Population Over Time](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/license vs population 2.png)
+
+  Although there is not an obvious and immediate relationship with our initial visualizations, it is still possible that we could find a relationship if we can find quick changes in the number of licenses in certain areas to exploit.
+Thus, in the following graph we explored the number of licenses in various states to see if any saw rapid increases or decreases in hunting licenses:
+![Hunting Licenses by State](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Hunting Licenses by State.png)
+
+From this visualization, we see that most states do not have significant variation in their number of hunting licenses.
+However, Texas saw an upward trend in licenses from 2012 to 2017.
+The following plot explores the relationship between the increase in hunting licenses and population changes in Texas:
+![License vs Population in Texas](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/License vs Population in Texas.png)
+
+The data does not seem to indicate a linear relationship.
+This could be due to a variety of reasons, including the possibility that an increase in the number of hunting licenses does not necessarily mean that there is an increase in the amount of Bison hunting (as the licenses could be for any number of animals).
+
+  The other state that saw some variation in the number of hunting licenses issued is Oklahoma, as shown below:
+![License vs Population in Texas](C:/Users/kdfro/Documents/MSU/GitHub/ECNS-560-Term-Project/Visuals/Oklahoma Hunting vs Pop.png)
+Although we still lack sufficient evidence to infer causality between the two variables, Oklahoma at least demonstrates a negative relationship between hunting licenses and Bison population.
+
+  We initially expected a negative relationship between hunting licenses and bison populations as a consequence of overhunting a common pool resource across all states.
+If states were to issue more licenses, we thought that this would ultimately mean that populations decrease as a consequence of more bison being killed by hunters.
+We believe that this relationship is not so obvious for two main reasons.
+First, the allocation of hunting licenses is heavily regulated at the state level in order to prevent a tragedy of the commons.
+Second, a state's capacity to maintain large Bison populations is dependent on the amount of land that the animals have to graze.
+Thus, an increase in the number of hunting licenses issued could in fact be an indicator of an increased amount of private hunting land.
+As the demand for Bison tags increases, it raises the opportunity cost of ranchers with land adjacent to bison ranges and incentives them to allow Bison to expand their territory.
+Thus, although more Bison are being hunted, there is more land available for additional growth.
+
+  In further exploration, other indicators such as weather or local policy differences may be more useful indicators of population levels.
+=======
 ---
@@
   In further exploration, other indicators such as weather or local policy differences may be more useful indicators of population levels.
+>>>>>>> ae5d3f30ef6b3c153d8c3a489f3f9dc3d1e461cb
